


# Appendix: Conventions for Notation  {#C:NotationConventionLC}

*Chapter Preview*. *Life Contingencies* employs widely accepted statistical concepts and tools. Thus, the notation should be consistent with standard usage employed in probability and mathematical statistics. See, for example, [@halperin1965recommended] for a description of one standard.


## General Conventions {#S:NC:General}


- Random variables are denoted by upper-case italicized Roman letters, with $X$ or $Y$ denoting a claim size variable, $N$ a claim count variable, and $S$ an aggregate loss variable. Realizations of random variables are denoted by corresponding lower-case italicized Roman letters, with $x$ or $y$ for claim sizes, $n$ for a claim count, and $s$ for an aggregate loss.
- Probability events are denoted by upper-case Roman letters, such as $\Pr(\mathrm{A})$ for the probability that an outcome in the event ''A'' occurs.
- Cumulative probability functions are denoted by $F(z)$ and probability density functions by the associated lower-case Roman letter: $f(z)$.
- For distributions, parameters are denoted by lower-case Greek letters. A caret or ''hat''  indicates a sample estimate of the corresponding population parameter. For example, $\hat{\beta}$ is an estimate of $\beta$ .
- The arithmetic mean of a set of numbers, say, $x_1, \ldots, x_n$, is usually denoted by $\bar{x}$; the use of $x$, of course, is optional.
- Use upper-case boldface Roman letters to denote a matrix other than a vector. Use lower-case boldface Roman letters to denote a (column) vector. Use a superscript prime ''$\prime$'' for transpose. For example, $\mathbf{x}^{\prime} \mathbf{A} \mathbf{x}$ is a quadratic form.
- Acronyms are to be used sparingly, given the international focus of our audience. Introduce acronyms commonly used in statistical nomenclature but limit the number of acronyms introduced. For example, *pdf* for probability density function is useful but *GS* for Gini statistic is not.



## Abbreviations   {#S:NC:Abbreviations}

Here is a list of abbreviations that we adopt. We italicize these acronyms. For example, we can discuss the goodness of fit in terms of the *AIC* criterion.

$$
\begin{array}{ll} \hline
\textbf{Symbol} & \textbf{Description} \\
\hline
AIC & \text{Akaike information criterion} \\
BIC & \text{(Schwarz) Bayesian information criterion} \\
cdf & \text{cumulative distribution function} \\
df & \text{degrees of freedom} \\
iid & \text{independent and identically distributed} \\
GLM & \text{generalized linear model} \\
mle & \text{maximum likelihood estimate/estimator}\\
ols & \text{ordinary least squares} \\
pdf & \text{probability density function} \\
pmf & \text{probability mass function} \\ \hline
\end{array}
$$


## Common Statistical Symbols and Operators  {#S:NC:StatSymbols}

Here is a list of commonly used statistical symbols and operators, including the latex code that we use to generate them (in the parens).

$$
\begin{array}{cl}  \hline
\textbf{Symbol} & \textbf{Description} \\
\hline
I(\cdot) & \text{binary indicator function (}I\text{). For example, }I(A) \text{ is one if an outcome in event} \\
& \ \ \ \ \  A \text{ occurs and is 0 otherwise.} \\
\Pr(\cdot) & \text{probability }(\backslash{\tt{Pr}}) \\
\mathrm{E}(\cdot)  & {\text{expectation operator }} (\backslash{\tt{mathrm\{E\}}}). {\text{ For example, }} \mathrm{E}(X)=\mathrm{E}~X {\text{ is the }} \\
& \ \ \ \ \ {\text{expected value of the random variable }}X,{\text{ commonly denoted by }}\mu. \\
\mathrm{Var}(\cdot)  & \text{variance operator }(\backslash{\tt{mathrm\{Var\}}}). \text{ For example, } \mathrm{Var}(X)=\mathrm{Var}~X\text{ is the} \\
& \ \ \ \ \  \text{ variance of the random variable } X, \text{commonly denoted by } \sigma^2. \\
\mu_k = \mathrm{E}~X^k & \text{kth moment of the random variable X. For }k\text{=1, use }\mu = \mu_1. \\
\mathrm{Cov}(\cdot,\cdot)  & \text{covariance operator } (\backslash{\tt{mathrm\{Cov\}}}).\text{ For example, } \\
& \ \ \ \ \ \mathrm{Cov}(X,Y)=\mathrm{E}\left\{(X -\mathrm{E}~X)(Y-\mathrm{E}~Y)\right\}  =\mathrm{E}(XY) -(\mathrm{E}~X)(\mathrm{E}~Y)\\
& \ \ \ \ \  \text{ is the covariance between random variables }X\text{ and }Y. \\
\mathrm{E}(X | \cdot)  & \text{conditional expectation operator. For example, }\mathrm{E}(X |Y=y) \text{ is the}\\
& \ \ \ \ \   \text{ conditional expected value of a random variable }X\text{ given that }\\
& \ \ \ \ \   \text{ the random variable }Y\text{ equals y. }\\
\Phi(\cdot) & \text{standard normal cumulative distribution function }(\backslash{\tt{Phi}})\\
\phi(\cdot) & \text{standard normal probability density function }(\backslash{\tt{phi}})\\
\sim & \text{means is distributed as }(\backslash{\tt{sim}}). \text{ For example, }X\sim F \text{ means that the } \\
& \ \ \ \ \  \text{random variable } X \text{ has distribution function }F. \\
se(\hat{\beta}) & \text{standard error of the parameter estimate }\hat{\beta} ~ (\backslash{\tt{hat\{}}\backslash{\tt{beta\}}}), \text{ usually }\\
& \ \ \ \ \  \text{ an estimate of the standard deviation of }\hat{\beta},\text{ which is }\sqrt{Var(\hat{\beta})}. \\
H_0 &  \text{null hypothesis} \\
H_a \text{ or }H_1 & \text{alternative hypothesis} \\
\hline
\end{array}
$$

  
  
## Common Mathematical Symbols and Functions  {#S:NC:Symbols}

Here is a list of commonly used mathematical symbols and functions, including the latex code that we use to generate them (in the parens).

$$
\begin{array}{cll}
\hline
\textbf{Symbol} & \textbf{Latex code} & \textbf{Description} \\
\hline
\equiv & \backslash\tt{equiv} & \text{identity, equivalence} \\
\implies  & \backslash\tt{implies} & \text{implies} \\
\iff  & \backslash\tt{iff} & \text{if and only if} \\
\to, \longrightarrow & \backslash\tt{to}, \backslash\tt{longrightarrow} & \text{converges to} \\
\mathbb{N} & \backslash\tt{mathbb\{N\}} & \text{natural numbers }1,2,\ldots \\
\mathbb{R} & \backslash\tt{mathbb\{R\}} & \text{real numbers} \\
\in        & \backslash\tt{in} & \text{belongs to} \\
\notin     & \backslash\tt{notin} & \text{does not belong to} \\
\subseteq  & \backslash\tt{subseteq} & \text{is a subset of} \\
\subset    & \backslash\tt{subset} & \text{is a proper subset of} \\
\cup       & \backslash\tt{cup} & \text{union} \\
\cap       & \backslash\tt{cap} & \text{intersection} \\
\emptyset  & \backslash\tt{emptyset} & \text{empty set} \\
A^{c}      &  & \text{complement of }A   \\
g*f        &  & \text{convolution }(g*f)(x)=\int_{-\infty}^{\infty}g(y)f(x-y)dy \\
\exp       & \backslash\tt{exp} & \text{exponential} \\
\log       & \backslash\tt{log} & \text{natural logarithm }\\
\log_a     &  & \text{logarithm to the base }a \\
!          &  & \text{factorial} \\
\text{sgn}(x) & \backslash\tt{sgn} & \text{sign of x} \\
\lfloor x\rfloor & \backslash\tt{lfloor}, \backslash\tt{rfloor} & \text{integer part of x, that is, largest integer }\leq x \\
|x|        &  & \text{absolute value of scalar }x \\
\varGamma(x) & \backslash\tt{varGamma} & \text{gamma (generalized factorial) function},\\
           &  & \text{satisfying }\varGamma(x+1)=x\varGamma(x) \\
B(x,y)     &  & \text{beta function, }\varGamma(x)\varGamma(y)/\varGamma(x+y) \\
\hline
\end{array}
$$

## Actuarial Notation  {#S:NC:ActSymbols}

Here is a list of commonly used actuarial symbols and functions, including the latex code that we use to generate them.

### Life Table Notation 

$$
{\small
\require{enclose}
\begin{array}{cll}  \hline
\textbf{Symbol} & \textbf{Latex code} & \textbf{Description} \\
\hline
\ell_{x} & \verb|\| \tt{ell}\_\{\tt{x}\} & \text{Expected number of lives at age } x \\
\ell_{[x]} & \backslash\tt{ell}\_\{ [\tt{x}]\} & \text{Expected number of lives at select age } x \\
_{t}p_{x} &    ~ \_\{ \tt{t} \} \tt{p} \_ \{\tt{x}\} & \text{Probability that a life aged } x \text{ survives }t \text{ years} \\
_{t}p_{[x]} &    ~ \_\{ \tt{t} \} \tt{p} \_ \{ [\tt{x}]\}\} & \text{Probability that a select life aged } x \text{ survives }t \text{ years} \\
\mathring{e}_{x} & \backslash\tt{mathring\{e\}} \_ \{ x\} & \text{Complete expectation of life at age } x \\
\mathring{e}_{x:{\enclose{actuarial}{n}}} & \backslash\tt{mathring\{e\}} \_ \{ x:\{\backslash enclose\{actuarial\}\{n\}\}\} & \text{Complete expectation of life at age } x \text{ for the next } n \text{ years} \\
\hline
\end{array}
}
$$

### Annuity Certain and Life Annuity Notation 

$$
{\small
\begin{array}{cll}  \hline
\textbf{Symbol} & \textbf{Latex code} & \textbf{Description} \\
\hline
a_{\enclose{actuarial}{n}} & \tt{a}\_\{\backslash \tt{enclose}\{actuarial\}\{\tt{n}\}\} & n\text{-year annuity certain-immediate} \\
\ddot{a}_{\enclose{actuarial}{n}} & \backslash\tt{ddot}\_\{\tt{a}\_\{\backslash \tt{enclose}\{actuarial\}\{\tt{n}\}\} & n\text{-year annuity certain-due} \\
\ddot{a}^{(m)}_{\enclose{actuarial}{n}} & \backslash\tt{ddot} \verb|^| \tt{\{(m)\}}\_\{\tt{a}\_\{\backslash \tt{enclose}\{actuarial\}\{\tt{n}\}\} & n\text{-year annuity certain-due payable } m\text{-thly}  \\
_{m|}\ddot{a}_{\enclose{actuarial}{n}} & ~ \_\{ \tt{m|} \} \backslash\tt{ddot}\_\{\tt{a}\_\{\backslash \tt{enclose}\{actuarial\}\{\tt{n}\}\} & m\text{-year deferred } n\text{-year annuity certain-due} \\
a_{x} & \tt{a}\_\{x\} & \text{APV of a whole life annuity immediate to } (x) \\
\ddot{a}_{x} & \backslash\tt{ddot}\{\tt{a}\}\_\{x\} & \text{APV of a whole life annuity due to } (x) \\
a^{(m)}_{x} & \tt{a} \verb|^| \tt{\{(m)\}}\_\{x\} & \text{APV of a life annuity immediate to } (x) \text{ payable } m\text{-thly} \\
\ddot{a}^{(m)}_{x} & \backslash\tt{ddot}\{\tt{a}\} \verb|^| \tt{\{(m)\}}\_\{x\} & \text{APV of a life annuity due to } (x) \text{ payable } m\text{-thly} \\
_{n|}\ddot{a}_{x} & ~ \_\{ \tt{n|} \} \backslash\tt{ddot}\_\{\tt{a}\_\{x\}\} & \text{APV of an } n\text{-year deferred life annuity due to } (x) \\
\ddot{a}_{x:{\enclose{actuarial}{n}}} & \backslash\tt{ddot\{a\}} \_ \{ x:\{\backslash enclose\{actuarial\}\{n\}\}\} & \text{APV of an } n\text{-year life annuity due to } (x) \\
\ddot{a}^{(m)}_{x:{\enclose{actuarial}{n}}} & \backslash\tt{ddot\{a\}} \verb|^| \tt{\{(m)\}} \_ \{ x:\{\backslash enclose\{actuarial\}\{n\}\}\} & \text{APV of an } n\text{-year life annuity due to } (x) \text{ payable } m\text{-thly} \\
\hline
\end{array}
}
$$

Not sure if necessary to say APV (actuarial present value)

### Life Insurance Notation 


$$
{\small
\begin{array}{cll}  \hline
\textbf{Symbol} & \textbf{Latex code} & \textbf{Description} \\
\hline
A_{x} &    \tt{A} \_ \{\tt{x}\} & \text{APV of a discrete whole life insurance to } (x) \\
\bar{A}_{x} &    \backslash \tt{bar} \{\tt{A}\} \_ \{\tt{x}\} & \text{APV of a continuous whole life insurance to } (x) \\
A^{(m)}_{x} &    \tt{A} \verb|^| \tt{\{(m)\}}\_ \{\tt{x}\} & \text{APV of a whole life insurance payable at end of } m\text{-th of the year of death of } (x) \\
_{n}E_{x} &    ~ \_\{ \tt{n} \} \tt{E} \_ \{\tt{x}\} & \text{APV of an } n\text{-year pure endowment to } (x) \\
{A}_{x:\enclose{actuarial}{n}}^{\quad 1} & \{\tt{A}\} \_ \{ x:\{\backslash enclose\{actuarial\}\{n\}\}\} \verb|^| \{\backslash \tt{quad} \ 1\} & \text{APV of an } n\text{-year term insurance to } (x)  \\
{A}_{x:{\enclose{actuarial}{n}}} & \{\tt{A}\} \_ \{ x:\{\backslash enclose\{actuarial\}\{n\}\}\} & \text{APV of an } n\text{-year endowment insurance to } (x) \\
_{n|}{A}_{x} & ~ \_\{ \tt{n|} \} \{\tt{A}\_\{x\}\} & \text{APV of an } n\text{-year deferred life insurance to } (x) \\
{A}_{x:\enclose{actuarial}{n}}^{\space 1} & \{\tt{A}\} \_ \{ x:\{\backslash enclose\{actuarial\}\{n\}\}\} \verb|^| \{\backslash \tt{space} \ 1\} & \text{APV of an } n\text{-year term insurance to } (x)  \\
\hline
\end{array}
}
$$

## Halo System  {#C:NotConv:Halo}


```{r  Halo1, warning=FALSE,message=FALSE, comment=NA, eval=FALSE, echo = FALSE}
plot.new()
par(mar = c(0,0,0,0) + 0.1)
plot.window(xlim=c(0,10),ylim=c(0,10))

# library(latex2exp)

text(0,10,1, cex = 1.5)
arrows(0.5,9.5,4,6,code=2,lwd=2,angle=45,length=0.10)
text(5,10,2, cex = 1.5)
arrows(  5,9.5,5,6,code=2,lwd=2,angle=45,length=0.10)
text(10,10,3, cex = 1.5)
arrows(9.5,9.5,6,6,code=2,lwd=2,angle=45,length=0.10)
text(10,0,4, cex = 1.5)
arrows(9.5,0.5,6.8,3.8,code=2,lwd=2,angle=45,length=0.10)
text(5,0,5, cex = 1.5)
arrows(  5,0.5,5,4,code=2,lwd=2,angle=45,length=0.10)
text(0,0,6, cex = 1.5)
arrows(0.5,0.5,4,4,code=2,lwd=2,angle=45,length=0.10)

text(5,5,expression(paste(symbol("\040")[paste(italic(m),"|")]^{~2},scriptstyle(symbol("\040")),bar(italic(A))[paste(italic(x),":",bar(italic(n)),"|")]^{scriptscriptstyle(1)})), cex=3)
 
```

</div>

(ref:Halo1a) **Halo System: Example of an Actuarial Symbol**. *Source*: This is adapted from the *Wikipedia* page [Actuarial Notation](https://en.wikipedia.org/wiki/Actuarial_notation).


```{r Halo1a, ref.label = 'Halo1', fig.cap = '(ref:Halo1a)', fig.align = 'center', out.width='50%', echo = FALSE}
```

***

\newpage

**Legend:**

*  1 - the "2" means double the force of interest
*  2 - the bar implies continuous, or payments at the moment of death. No bar, or blank, signifies payment at the end of the year.
For annuities, could use a double dot that signifies payments at the beginning of the year.
*  3 - the "1" denotes a temporary, or term, payment. Paid if $x$ within $n$ years
*  4 - contract issued to a life aged $x$ and lasts for at most $n$ years.
*  5 - the "A" signifies it is an insurance ("assurance") contract. Could also be a lower case $a$ that would signify an annuity contract.
*  6 - deferred *m* years



**Note:** It is sometimes helpful to put a space before the pre-subscript like $\frac{\ell_{[x]+t}}{\ell_{[x]}} = ~_{t}p_{[x]}$. Else you get this awkward $\frac{\ell_{[x]+t}}{\ell_{[x]}} = _{t}p_{[x]}$.


$$
Y =
 \begin{cases}
 1 + v + v^2 + \cdots + v^K = \ddot{a}_{\enclose{actuarial}{K+1}}, K=0,1,\ldots,n-1 \\
 1 + v + v^2 + \cdots + v^{n-1} = \ddot{a}_{\enclose{actuarial}{n}}, K \ge n
 \end{cases}
$$
**Note.**

*  For the `html` version of the text, we are using the [STACK]() system; in particular, see their [actuarial notation](https://stack-demo.maths.ed.ac.uk/demo/question/type/stack/doc/doc.php/Authoring/Actuarial.md) page.
*  This does not appear to work for the `pdf` version, at least with R-Studio. The latex package [actuarialangle](https://ctan.org/pkg/actuarialsymbol?lang=en) provides a substitute. To make this happen, in our latex header .tex page, we have the following bit of code

```
%  Something for the Actuarial Angle package
\usepackage{actuarialangle}
\newcommand{\enclose}[1]{\actuarialangle} 
```


## Further Readings

To make connections to other literatures, see [@abadir2002notation] \url{http://www.janmagnus.nl/misc/notation.zip} for a summary of notation from the econometrics perspective. This reference has a terrific feature that many latex symbols are defined in the article. Further, there is a long history of discussion and debate surrounding actuarial notation; see [@boehm1975thoughts] for one contribution.



